---
title: "LDA on Zymbo data"
author: "Pratheepa Jeganathan, Susan Holmes"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  BiocStyle::html_document
params:
  K: 4
  R: 500
---

# Outline

We have 10 negative control samples and serial dilution of ZymoBIOMICS® Microbial Community Standard.

Looking at the DADA2 pipeline ASVs 2, 3, 4, 5, 6, 7, 8, 11, 17, 37, 22 are in the ZymoBIOMICS® Microbial Community Standard corresponding to eight bacterial species. We used BARBI to identify DNA contaminants. 

We provide a latent Dirichlet allocation (LDA) to identify bacterial communities related to contaminant taxa. This modeling helps to identify contaminant sources associated with each bacterial community. 


# Set-up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
  message = FALSE, 
  warning = FALSE,
  fig.width = 12,
  fig.height = 8)
```

```{r read_arg}
K = params$K
K
iter = params$R
iter
```


Loading packages
```{r}
library(phyloseq)
library(dplyr)
library(magrittr)
library(rstan)
library(RColorBrewer)
library(wesanderson)
library(ggheatmap)# Github repo: #devtools::install_github("jakobbossek/ggheatmap")
library(reshape2)
library(DESeq2)
#library(Rcpp)
library(ape)
```

# Analysis

## Read phyloseq and exploratory analysis

```{r}
ps = readRDS("./Data/ps_zymo.rds")
if(dim(otu_table(ps))[1]!=ntaxa(ps)){
  otu_table(ps) = t(otu_table(ps))
  }
```

## Change order of sample names
```{r}
ncont = paste0("NegativeControl.", seq(1, 10))
stan = paste0("Standard.Dilution.1.",c(1, 6, 36, 216, 1296, 7776, 46656, 279936))

sample_data(ps)$Name = factor(sample_data(ps)$Name, levels = c(ncont,stan))

sample_names(ps) = as.character(sample_data(ps)$Name)
```

## Remove unequal library size depth effect
```{r}
ot = otu_table(ps) %>% data.frame()
geo_mean = function(x) {
        if(all(x == 0)){
            val = 0
        }else{
            val = exp(sum(log(x[x > 0]))/length(x))
        }
        return(val)
    }

geom_mean_row = apply(ot, 1, FUN = geo_mean)


dj = estimateSizeFactorsForMatrix(ot, median, geoMeans = geom_mean_row)

ot = t(t(ot)/dj)

ot = apply(ot , c(1,2), function(y){ as.integer(y)})
colnames(ot) = sample_names(ps)

ps = phyloseq(otu_table(ot, taxa_are_rows = TRUE), sample_data(ps), tax_table(ps))
ps
```


## Adding blocks

We have only one block of samples.

```{r adding_blocks}
blocks = rep("Set1", nsamples(ps))

sample_data(ps)$block = blocks
```

Identify the species that are not present in at least one `SampleType == Standard` sample and removed them from the phyloseq object. Label these species as contaminants. 

```{r filter_taxa}
ps = prune_taxa(taxa_sums(ps) > 0, ps)
ps.standard = subset_samples(ps, SampleType %in% c("Standard"))
prevTaxaP = apply(otu_table(ps.standard), 1, function(x){sum(x>0)})

Contaminants1 = names(prevTaxaP)[prevTaxaP == 0]
length(Contaminants1)
```

```{r}
ps = prune_taxa(prevTaxaP > 0, ps)
ps
```

We identified 142 ASVs not is any dilution series samples and they are classified as contaminants before using LDA.
  
See a summary of the `Standard` and `Negative` samples in each block. 

```{r summary_stat}
table(sample_data(ps)$SampleType, sample_data(ps)$block)
```

```{r}
colSums(otu_table(ps))
```

##  Prepare the phyloseq object for the Bayesian inference

Zymo does not have batch effect. 

If there is bacth effect, we split the phyloseq object into multiple phyloseq objects corresponding to each block, and store the phyloseq objects as a list of phyloseq objects, `psByBlock`. 

Select negative control samples from each block and store as a list of phyloseq objects, `psNCbyBlock`. 

Select all species that have a prevalence of zero (i.e., have zero reads) in all negative control samples for each block and store as a list of phyloseq objects, `psallzeroInNC`.

Select all plasma samples from each block and store as a list of phyloseq objects, `psPlByBlock`.

```{r list_of_phyloseq}
source("psBlockResults.R")
psBlockResult = psBlockResults(ps, 
                               sampleTypeVar = "SampleType",
                               caselevels = c("Standard"),
                               controllevel="Negative",
                               sampleName = "Name", 
                               blockVar = "block")

psByBlock = psBlockResult[[1]]
psNCbyBlock = psBlockResult[[2]]
psallzeroInNC = psBlockResult[[3]]
psPlByBlock = psBlockResult[[4]]

```

## LDA


### LDA on dilution series samples and negative controls

We choose the number of topics to be 4.

Read stan model
```{r}
f <- stan_model(file = "lda.stan")
```

Make stan data
```{r}
K = 4 # number of topics
ps = psByBlock[[1]] 
short.sample.names = c(paste0("NC.", c(1,10)), 
                       paste0("Di.", seq(1,8)), 
                       paste0("NC.", seq(2,9)))
sample_names(ps) = short.sample.names

x = t(get_taxa(ps))
dimnames(x) = NULL
stan.data <- list(K = K, 
  V = ncol(x), 
  D = nrow(x), 
  n = x, 
  alpha = rep(1, K), 
  gamma = rep(0.5, ncol(x))
)
```

```{r}
t1 = proc.time()
stan.fit = sampling(f,
  data = stan.data,
  iter = iter,
  chains = 1,
  sample_file = "./Results/sample_file_Zymo",
  diagnostic_file = "./Results/diagnostic_file_Zymo",
  cores = 1)

proc.time() - t1

saveRDS(stan.fit, file = "./Results/Zymo_LDA.rds")

# stan.fit = vb(f,
#   data = stan.data,
#   output_samples = 100,
#   eta = 1,
#   adapt_engaged = FALSE)
# saveRDS(stan.fit, file = "./Results/Zymo_LDA.rds")
```

```{r}
stan.fit = readRDS("./Results/Zymo_LDA.rds")
```

### Extract posterior samples 
- $\beta$ and $\theta$
```{r}
samples = rstan::extract(stan.fit, permuted = TRUE, inc_warmup = FALSE, include = TRUE)
```

### Posterior distribution of topic in each sample 
- $\theta$ for each sample
```{r}
theta = samples$theta # chain * sample * topic
# find theta distirbution of each topic in each sample
names(theta) = c("theta", "Sample", "Topic")
dimnames(theta)[[2]] = sample_names(ps)
dimnames(theta)[[3]] = c(paste0("Topic ", seq(1,K)))

theta.all = melt(theta)
colnames(theta.all) = c("Chain", "Sample", "Topic", "topic.dis")

theta.all$Sample = factor(theta.all$Sample)
theta.all$Topic = factor(theta.all$Topic)

# add control or dilution series
sam = sample_data(ps) %>% data.frame()
sam$unique_names = rownames(sam)
theta.all = left_join(theta.all, sam, by =c("Sample"= "unique_names"))

# theta.all =  mutate(theta.all, Group = ifelse(substr(Sample,1,1)=="D","Dilution", "Control"))

theta.all$Sample_Type = factor(theta.all$Sample_Type)
```


### boxplot

topic distirbution in each sample
```{r}
p = ggplot(data = theta.all) + 
  geom_boxplot(aes(x = Sample, y = topic.dis, color = Topic)) + 
  facet_grid(Topic ~ Sample_Type, scales = "free_x") +
  ylab(bquote(theta[k])) + 
  ggtitle("") + 
  theme(plot.title = element_text(hjust = 0.5, size = 15), 
    strip.text.y= element_text(size = 12), 
    strip.text.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    legend.position = "none")
p

ggsave("./Figures/Zymo_LDA_theta_col.eps", plot = p, width = 10, height = 7)
```


We chose four topics because two topics always dominated in dilution series samples. 
<!-- When we chose three topics, one of the topics dominated in dilution series sample is also dominated in control sample. Topic 1 is contaminant topic because it's only dominated in control samples. Topic 3 is also a contaminant topic because $\theta$ distribution range is less than .25 in all dilution samples. Topic 2 is real species topic because it is highly dominated in dilution series samples. Topic 4 is mix of mostly real and some contribution from contaminant species. -->

### ASV cloud in each topic
```{r}
beta = samples$beta
dimnames(beta)[[2]] = c(paste0("Topic ", seq(1,K)))

tax_tab = tax_table(ps) %>% data.frame()
tax_tab = mutate(tax_tab, seq.variant = rownames(tax_tab))

dimnames(beta)[[3]] =tax_tab[, "seq.variant"]
beta.all = melt(beta)
colnames(beta.all) = c("Chain", "Topic", "ASV", "ASV.distribution")
beta.all = left_join(beta.all, tax_tab, by = c("ASV"= "seq.variant"))
beta.all$Topic = factor(beta.all$Topic)
beta.all$ASV = factor(beta.all$ASV)
```

compute max of $\beta_{k}$ for each ASV in each topic

```{r}
max.beta.in.each.asv.all.topics = group_by(beta.all, Topic, Family, Genus) %>% summarise(max_beta = max(ASV.distribution)) %>% top_n(10, max_beta) %>% as.data.frame()


library(ggwordcloud)

p_cloud = ggplot(max.beta.in.each.asv.all.topics, 
                 aes(label = Genus, size = max_beta, color = Family)) + 
  geom_text_wordcloud() +
  theme_minimal() +
  scale_size_area(max_size = 11) + 
  facet_wrap(~ Topic) + 
  theme(strip.text.x = element_text(size = 12, face = "bold"))

p_cloud

ggsave("./Figures/Zymo_ASV.png", p_cloud, width = 10, height = 7, units = "in", dpi = 300)
```


