---
title: "LDA-VB (Variational Bayes) on ED data"
author: "Pratheepa Jeganathan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
  message = FALSE, 
  warning = FALSE, 
  fig.width = 16, 
  fig.height = 12)
```

Loading packages
```{r}
library(phyloseq)
library(ggplot2)
library(magrittr)
library(dplyr)
library(reshape2)
library(rstan)
library(RColorBrewer)
library(wesanderson)
library(ggheatmap)
```


### Read phyloseq and exploratory analysis

```{r eval=FALSE, include=FALSE}
ps = readRDS("./Data/ps_sirs_updated_may_1_2018.rds")
ps
if(dim(otu_table(ps))[1]!=ntaxa(ps)){otu_table(ps) <- t(otu_table(ps))}
ps = subset_samples(ps, !(Sample_Type == "Library_Control"))
mapp = readxl::read_excel("./Data/Pt_and_P.xlsx")
sample_names(ps)[which(sample_names(ps) %in% mapp$P)] = mapp$Pt
sample_data(ps)$SampleCode = sample_names(ps)

ps = subset_samples(ps, !(SampleType == "Healthy"))
saveRDS(ps, "./Data/ps_ED.rds")
```

```{r}
ps = readRDS("./Data/ps_ED.rds")

set1 = c("1","2","3","4","11","12")
set2 = c("5","6","7","8","9","10")
setP = "P"
ext.num = sample_data(ps)$Extraction_Number

blocks = ifelse(ext.num %in% set1, "Set1", ifelse(ext.num %in% set2, "Set2","SetP"))

sample_data(ps)$block = blocks
sample_data(ps)$block = factor(sample_data(ps)$block)

if(dim(otu_table(ps))[1]!=ntaxa(ps)){
  otu_table(ps) = t(otu_table(ps))
}

```

Identify the species that are not present in at least one `Sample_Type == Plasma` sample and removed them from the phyloseq object. Label these species as contaminants. 

```{r filter_taxa}
ps = prune_taxa(taxa_sums(ps) > 0, ps)
ps.standard = subset_samples(ps, Sample_Type %in% c("Plasma"))
prevTaxaP = apply(otu_table(ps.standard), 1, function(x){sum(x>0)})

Contaminants1 = names(prevTaxaP)[prevTaxaP == 0]
length(Contaminants1)
```

```{r}
ps = prune_taxa(prevTaxaP > 0, ps)
ps
```

We identifed 369 species not in any plasma samples and they are classified as contaminants before using BARBI LDA (VB).
  
See a summary of the `plasma` and `control` samples in each block. 

```{r summary_stat}
table(sample_data(ps)$Sample_Type, sample_data(ps)$block)
```


###      Prepare the phyloseq object for the Bayesian inference

We use Bayesian inference to identify contaminants in each block separately to account for the batch-effects of contamination. 

Thus, split the phyloseq object into multiple phyloseq objects corresponding to each block, and store the phyloseq objects as a list of phyloseq objects, `psByBlock`. 

Select negative control samples from each block and store as a list of phyloseq objects, `psNCbyBlock`. 

Select all species that have a prevalence of zero (i.e., have zero reads) in all negative control samples for each block and store as a list of phyloseq objects, `psallzeroInNC`.

Select all plasma samples from each block and store as a list of phyloseq objects, `psPlByBlock`.

```{r list_of_phyloseq}
source("psBlockResults.R")
psBlockResult = psBlockResults(ps, 
  sampleTypeVar = "Sample_Type", caselevels = c("Plasma"), controllevel="Control", sampleName = "SampleCode", blockVar = "block")

psByBlock = psBlockResult[[1]]
psNCbyBlock = psBlockResult[[2]]
psallzeroInNC = psBlockResult[[3]]
psPlByBlock = psBlockResult[[4]]
```

## LDA using VB

Each sample is modeled as having a mixture of topics, with each species drawn from a topic based on mixing proportions.


### LDA on specimen and control samples


```{r eval=FALSE}
f = stan_model(file = "lda.stan")
```

For each block, make a stan data.

```{r}
K = as.list(c(5,4,3)) # number of topics
stan.data.all.blocks = lapply(as.list(1:3), function(i){
  y = psByBlock[[i]]
  x = t(get_taxa(y))
  dimnames(x) = NULL
  stan_data = list(K = K[[i]], 
    V = ncol(x), 
    D = nrow(x), 
    n = x, 
    alpha = rep(1, K[[i]]), 
    gamma = rep(0.5, ncol(x)))
  return(stan_data)
})
```

Sample from posterior using variational Bayes.
```{r eval=FALSE}
stan.fit.all.blocks = lapply(stan.data.all.blocks, function(u){
  stan_fit = vb(f, 
    data = u, 
    output_samples = 1000, 
    eta = 1, 
    adapt_engaged = FALSE,
    seed = 234
)
  return(stan_fit)
})


saveRDS(stan.fit.all.blocks, file = "./Results/stan_fit_all_ED.rds")
```

```{r}
stan.fit.all.blocks <- readRDS("./Results/stan_fit_all_ED.rds")
```

### Extract posterior samples
```{r}
samples.all.blocks <- lapply(stan.fit.all.blocks, function(x){
  rstan::extract(x)
})
```

### Posterior distirbution of topic in each sample in each block
```{r}
theta.all.blocks = lapply(as.list(1:3), function(i){
  x = samples.all.blocks[[i]]
  theta = x$theta # chain * sample * topic
  # find theta distirbution of each topic in each sample
  names(theta) = c("theta", "Sample", "Topic")
  theta.all = melt(theta)
  colnames(theta.all) = c("Chain", "Sample", "Topic", "topic.dis")
  theta.all$Sample = factor(theta.all$Sample)
  #levels(theta.all$Sample) = sample_names(ps)
  theta.all$Topic = factor(theta.all$Topic)
  levels(theta.all$Topic) = c(paste0("Topic_", seq(1,K[[i]])))
  return(theta.all)
})

for(i in 1:3){
  levels(theta.all.blocks[[i]]$Sample) = sample_names(psByBlock[[i]])
  # add control or plasam
  theta.all.blocks[[i]] =  mutate(theta.all.blocks[[i]], Group = ifelse(substr(Sample, 1, 2)=="Pt","Plasma", "Control"))
  theta.all.blocks[[i]]$Group = factor(theta.all.blocks[[i]]$Group)
}
```

### Plot topic distribution in each sample in each block
```{r}
col.topic = c(wes_palette("Cavalcanti1", n = 5), wes_palette("Darjeeling1", n = 5)) # sum of two n's should be equal to K.

p.all.blocks = lapply(theta.all.blocks, function(u){
  #sel.samples  = unique(u$Sample)[1:20]
  sel.samples = c(as.character(unique(u$Sample[(u$Group == "Control")])), as.character(unique(u$Sample[(u$Group == "Plasma")])[c(1:10, 48)]))
  u = filter(u, Sample %in% sel.samples)
  #u = filter(u, Sample %in% c(sample_names(psByBlock[[1]])[1:20])) # this is to make only few sample plots
  # p = ggplot() + 
  #   geom_density(data = u, 
  #     aes(x = topic.dis, 
  #       y = ..scaled.., 
  #       col = Topic, 
  #       fill=Topic)) + 
  #   facet_grid(Sample ~ Group, 
  #     scales = "fixed") + 
  #   scale_color_manual(values = col.topic) + 
  #   scale_fill_manual(values = col.topic) + 
  #   ylab("density") +
  #   xlab("") + 
  #   ggtitle("") + 
  #   theme(plot.title = element_text(hjust = 0.5), 
  #     strip.text.y= element_text(size = 12, angle = 360), 
  #     strip.text.x = element_text(size  = 12))
  
    p = ggplot() + 
    geom_density(data = u, 
      aes(x = topic.dis, 
        y = ..scaled.., 
        linetype = Topic)) + 
    facet_grid(Sample ~ Group, 
      scales = "fixed") + 
    ylab("density") +
    xlab("") + 
    ggtitle("") + 
    theme(plot.title = element_text(hjust = 0.5), 
      strip.text.y= element_text(size = 12, angle = 360), 
      strip.text.x = element_text(size  = 12))
  return(p)
})

```

```{r}
p.all.blocks[[1]]
ggsave("./Figures/LDA_ED_set1_2.eps", plot = p.all.blocks[[1]], width = 12, height = 16)
```

```{r}
p.all.blocks[[2]]
ggsave("./Figures/LDA_ED_set2.eps", plot = p.all.blocks[[2]], width = 12, height = 16)
```

```{r}
p.all.blocks[[3]]
ggsave("./Figures/LDA_ED_setP.eps", plot = p.all.blocks[[3]], width = 12, height = 16)
```


### boxplot as in Kris's paper
```{r}

p.all.blocks = lapply(theta.all.blocks, function(u){
  sel.samples = c(as.character(unique(u$Sample[(u$Group == "Control")])), as.character(unique(u$Sample[(u$Group == "Plasma")])[c(1:10, 48)]))
  u = filter(u, Sample %in% sel.samples)
  p = ggplot(data = u) + 
  geom_boxplot(aes(x = Sample, y = topic.dis, color = Topic)) + 
  facet_grid(Topic ~ Group, scales = "free_x")+
  ylab(bquote(theta[k])) + 
  ggtitle("") + 
  theme(plot.title = element_text(hjust = 0.5, size = 15), 
    strip.text.y= element_text(size = 6), 
    strip.text.x = element_text(size = 6) , axis.text.x = element_text(angle = 90, hjust = 1))
  
  return(p)
})

```

```{r}
p.all.blocks[[1]]
ggsave("./Figures/LDA_ED_set1_boxplot_col.eps", plot = p.all.blocks[[1]], width = 9, height = 6)
```


```{r}
p.all.blocks[[2]]
ggsave("./Figures/LDA_ED_set2_boxplot_col.eps", plot = p.all.blocks[[2]], width = 9, height = 6)
```

```{r}
p.all.blocks[[3]]
ggsave("./Figures/LDA_ED_set3_boxplot_col.eps", plot = p.all.blocks[[3]], width = 9, height = 6)
```

### Posterior ASV distribution in each topic in each block
```{r}
beta.all.blocks = lapply(as.list(1:3), function(i){
  x = samples.all.blocks[[i]]
  beta = x$beta
  beta.all = melt(beta)
  colnames(beta.all) = c("Chain", "Topic", "ASV", "ASV.distribution")
  beta.all$Topic = factor(beta.all$Topic)
  levels(beta.all$Topic) = c(paste0("Topic_", seq(1,K[[i]])))
  beta.all$ASV = factor(beta.all$ASV)
  return(beta.all)
})

for(i in 1:3){
  levels(beta.all.blocks[[i]]$ASV) = taxa_names(psByBlock[[i]])
}

```


```{r}
library(viridis)
heatmap.map.theta.blocks = lapply(beta.all.blocks, function(x){
  #sel.spec = as.character(unique(x$ASV)[1:50])
  #x = filter(x, ASV %in% sel.spec)
  x.summary = x %>% group_by(Topic, ASV) %>% summarise(log_median_beta = log(median(ASV.distribution, 
    na.rm = TRUE))) %>%
  as.data.frame()
  
  # choose species with higher proportions
  x.summary = x.summary[order(x.summary$log_median_beta, decreasing = TRUE),]
  x.summary = x.summary[1:50, ]
  
  p = ggheatmap(x.summary,
  id.vars = c("ASV", "Topic"),
  value.name = "log_median_beta"
) + 
    ggtitle("ASV distribution in each topic") + 
  theme(plot.title = element_text(hjust = 0.5), 
    axis.text.x = element_text(size = 12), 
    axis.text.y = element_text(size = 12)) +
  scale_fill_gradient(low = "white", high = "orange")
  return(p)
})

```

```{r}
heatmap.map.theta.blocks[[1]]
```

```{r}
heatmap.map.theta.blocks[[2]]
```

```{r}
heatmap.map.theta.blocks[[3]]
```

