---
title: "decontam"
author: "PJ, HC, SH, DR"
date: '`r format(Sys.time(), "%B %d, %Y")`'
header-includes:
  - \usepackage{color}
output: 
    html_document:
        toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(decontam)
library(dplyr)
library(magrittr)
library(phyloseq)
library(ggplot2)
```

# Use `decontam` on Zymo data


### Read phyloseq 

```{r}
ps <- readRDS("./Data/ps_zymo.rds")
if(dim(otu_table(ps))[1]!=ntaxa(ps)){otu_table(ps) <- t(otu_table(ps))}
```


### Change order of sample names
```{r}
ncont <- paste0("NegativeControl.",seq(1,10))
stan <- paste0("Standard.Dilution.1.", c(1, 6, 36, 216, 1296, 7776, 46656, 279936))
#stan <- c(paste0("Standard.Dilution.1.",c(1,6,36,216,1296,7776,46656)),"StandardDilution.1.279936")

sample_data(ps)$Name <- factor(sample_data(ps)$Name, levels = c(ncont,stan))

sample_names(ps) <- as.character(sample_data(ps)$Name)
```


### Store sequence variant, ASVs, ASV.genus, ASV.genus.species
```{r}
ASV <- as.character(paste0("ASV_",seq(1,ntaxa(ps))))
ASV.Genus <- paste0("ASV_",seq(1,ntaxa(ps)), "_", as.character(tax_table(ps)[,6]))
ASV.Genus.Species <- paste0(ASV,"_",as.character(tax_table(ps)[,6]), "_",  as.character(tax_table(ps)[,7]))

df.ASV <- data.frame(seq.variant = taxa_names(ps), ASV = ASV, ASV.Genus = ASV.Genus, ASV.Genus.Species = ASV.Genus.Species)
```


```{r}
taxa_names(ps) <- df.ASV$ASV.Genus.Species
```

### Microbial composition of the ZymoBIOMICS® Microbial Community Standard (Zymo Research, Irvine, CA) measured 

Top 20 ASVs in both control and dilution series samples

```{r}
top20 <- names(sort(taxa_sums(ps), decreasing=TRUE))[1:20]
ps.top20 <- transform_sample_counts(ps, function(OTU) OTU/sum(OTU))
ps.top20 <- prune_taxa(top20, ps.top20)
p <- plot_bar(ps.top20, x="Name", fill="Genus") + 
  facet_wrap(~SampleType, scales="free_x") + 
  theme_bw() + ylab("Relative abundance")
p
```

### True ASVs at genus level

```{r}
# true.com.ASV <- taxa_names(ps)[which(tax_table(ps)[,6] %in% c("Bacillus","Listeria","Staphylococcus", "Lactobacillus", "Escherichia/Shigella", "Enterococcus", "Salmonella", "Pseudomonas"))]
true.com.ASV <- taxa_names(ps)[c(2:8,11,17,37,22)]
ps.true <- transform_sample_counts(ps, function(OTU) OTU/sum(OTU))
ps.true <- prune_taxa(true.com.ASV, ps.true)
p <- plot_bar(ps.true, x="Name", fill="Genus") + 
  facet_wrap(~SampleType, scales="free_x") +
  theme_bw() + ylab("Relative abundance")
p
```

### We have only one block of samples.

```{r}
blocks <- rep("Set1", nsamples(ps))

sample_data(ps)$block <- blocks
```

### Remove ASV not in any standard samples
Identify the species that are not present in at least one `SampleType == Standard` sample and removed them from the phyloseq object. Label these species as contaminants. 

```{r}
ps <- prune_taxa(taxa_sums(ps) > 0, ps)
ps.standard <- subset_samples(ps, SampleType %in% c("Standard"))
prevTaxaP <- apply(otu_table(ps.standard), 1, function(x){sum(x>0)})

Contaminants1 <- names(prevTaxaP)[prevTaxaP == 0]
length(Contaminants1)
ps <- prune_taxa(prevTaxaP > 0, ps)
ps
```

We identified 142 ASVs not is any dilution series samples, and they are classified as contaminants before using BARBI.

Let's use decontam to infer remove contaminant ASVs in dilution series sample.

### Adding True Sample/Control Sample
```{r}
sam <- sample_data(ps) %>% data.frame()
sam <- mutate(sam, Sample_or_Control = ifelse(SampleType == "Standard", "True Sample", "Control Sample"))
sample_data(ps)$Sample_or_Control = sam$Sample_or_Control
```

### Inspect library sizes
```{r}
df <- as.data.frame(sample_data(ps)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Sample_or_Control)) + geom_point()
```

The library depth is not particularly large in Standard dilution samples than negative controls.

### Identify Contaminants - Frequency
We cannot used this method because DNA concentration is not available - not available

### Identify Contaminants - Prevalence

The second contaminant identification method we’ll use is the “prevalence” method. In this method, the prevalence (presence/absence across samples) of each sequence feature in true positive samples is compared to the prevalence in negative controls to identify contaminants.

In our phyloseq object, "Sample_or_Control" is the sample variable that holds the negative control information. We’ll summarize that data as a logical variable, with TRUE for control samples, as that is the form required by isContaminant.

```{r}
sample_data(ps)$is.neg <- sample_data(ps)$Sample_or_Control == "Control Sample"
contamdf.prev <- isContaminant(ps, method = "prevalence",
                               neg="is.neg")
table(contamdf.prev$contaminant)
```

We only identified only 6 contaminants.

```{r}
tax_table(ps)[which(contamdf.prev$contaminant), "Species"]
```

To remove more contaminant, we can use different threshold

### Change the threshold
In the prevalence test there is a special value worth knowing, threshold=0.5, that will identify as contaminants all sequences thare are more prevalent in negative controls than in positive samples. Let’s try using this more aggressive classification threshold rather than the default.


```{r}
contamdf.prev05 <- isContaminant(ps, method="prevalence",
                                 neg="is.neg", 
                                 threshold=0.5)
table(contamdf.prev05$contaminant)
```

We idetified 13 contaminants. Let's see what are they.

```{r}
tax_table(ps)[which(contamdf.prev05$contaminant), "Species"]

```

decontam identify more contaminants with threshold .5. 

### Prevalence of taxa in negative and standard dilution samples

Let’s take a look at the number of times several of these taxa were observed in negative controls and positive samples.
```{r}
# Make phyloseq object of presence-absence in negative controls and true samples
ps.pa <- transform_sample_counts(ps, function(abund) 1*(abund>0))
ps.pa.neg <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "Control Sample", ps.pa)
ps.pa.pos <- prune_samples(sample_data(ps.pa)$Sample_or_Control == "True Sample", ps.pa)
# Make data.frame of prevalence in positive and negative samples
df.pa <- data.frame(pa.pos=taxa_sums(ps.pa.pos), pa.neg=taxa_sums(ps.pa.neg),
                      contaminant=contamdf.prev$contaminant)
ggplot(data=df.pa, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)")
```

deontam still needs to identify more contaminant. When we increas threshold, decontam identifies increasing number of contaminants, but we are not sure which is the correct threshold.

# Use `decontam` on ED data

### Read phyloseq object 
```{r}
ps <- readRDS("./Data/ps_sirs_updated_may_1_2018.rds")
ps
if(dim(otu_table(ps))[1]!=ntaxa(ps)){otu_table(ps) <- t(otu_table(ps))}

ps <- subset_samples(ps, !(Sample_Type == "Library_Control"))
```

### Replace labels and remove healthy subjects
```{r}
mapp <- readxl::read_excel("./Data/Pt_and_P.xlsx")
sample_names(ps)[which(sample_names(ps) %in% mapp$P)] <- mapp$Pt
sample_data(ps)$SampleCode <- sample_names(ps)

ps <- subset_samples(ps, !(SampleType == "Healthy"))
```

### Identify batches
```{r}
set1 <- c("1","2","3","4","11","12")
set2 <- c("5","6","7","8","9","10")
setP <- "P"
ext.num <- sample_data(ps)$Extraction_Number

blocks <- ifelse(ext.num %in% set1, "Set1", ifelse(ext.num %in% set2, "Set2","SetP"))

sample_data(ps)$block <- blocks
```

### Remove healthy subjects

Subset samples such that only patient plasma and negative control samples are included in the final phyloseq object. 

```{r choose_samples}
ps <- subset_samples(ps,Sample_Type %in% c("Plasma","Control"))
```

###  Preprocessing

Identify the species that are not present in at least one plasma sample and removed those species from the phyloseq object. Label these species as contaminants. 

```{r}
ps <- prune_taxa(taxa_sums(ps) > 0, ps)
ps_plasma <- subset_samples(ps,Sample_Type %in% c("Plasma"))
prevTaxaP <- apply(otu_table(ps_plasma), 1 ,function(x){sum(x>0)})

Contaminants1 <- names(prevTaxaP)[prevTaxaP==0]
ps <- prune_taxa(prevTaxaP > 0,ps)
ps
```

### Adding True Sample/Control Sample
```{r}
sam <- sample_data(ps) %>% data.frame()
sam <- mutate(sam, Sample_or_Control = ifelse(SampleType == "SIRS", "True Sample", "Control Sample"))
sample_data(ps)$Sample_or_Control = sam$Sample_or_Control
```

### Inspect library sizes
```{r}
df <- as.data.frame(sample_data(ps)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data=df, aes(x=Index, y=LibrarySize, color=Sample_or_Control)) + geom_point()
```


The library sizes of the plasma samples primarily fall from 2 to 5000 reads and there are some high-read outliers. The negative control samples have large  number of reads (this contrasts with the expectation stated in decontam package). 

### Identify samples by blocks
```{r}
source("psBlockResults.R")
psBlockResult <- psBlockResults(ps, sampleTypeVar = "Sample_Type", caselevels = c("Plasma"), controllevel= "Control", sampleName = "SampleCode", blockVar = "block")

psByBlock <- psBlockResult[[1]]
```

### Identify Contaminants - Frequency
In this method, the distribution of the frequency of each sequence feature as a function of the input DNA concentration is used to identify contaminants.

In our phyloseq object, "Conc.ng.ul" is the sample variable that holds the concentration information:
```{r}
contamdf.freq.all.blocks = lapply(psByBlock, function(x){
  isContaminant(x, method="frequency", conc="Conc.ng.ul")
})
# contamdf.freq <- isContaminant(ps, method="frequency", conc="Conc.ng.ul")
head(contamdf.freq.all.blocks[[1]])
head(contamdf.freq.all.blocks[[2]])
head(contamdf.freq.all.blocks[[3]])
```

```{r}
table(contamdf.freq.all.blocks[[1]]$contaminant)
table(contamdf.freq.all.blocks[[2]]$contaminant)
table(contamdf.freq.all.blocks[[3]]$contaminant)
```

decontam identified only 1 species and 2 species as contaminants in block 1 and 2, respectively. 

```{r}
head(which(contamdf.freq.all.blocks[[1]]$contaminant))
head(which(contamdf.freq.all.blocks[[2]]$contaminant))
head(which(contamdf.freq.all.blocks[[3]]$contaminant))
```

```{r}
plot_frequency(psByBlock[[1]], taxa_names(ps)[c(1,3)], conc="Conc.ng.ul") +
  xlab("DNA Concentration (PicoGreen fluorescent intensity)")
```


### Identify Contaminants - Prevalence

The second contaminant identification method we’ll use the “prevalence” method. In this method, the prevalence (presence/absence across samples) of each sequence feature in true positive samples is compared to the prevalence in negative controls to identify contaminants.

In our phyloseq object, "Sample_or_Control" is the sample variable that holds the negative control information. We’ll summarize that data as a logical variable, with TRUE for control samples, as that is the form required by isContaminant.

```{r}
psByBlock = lapply(psByBlock, function(x){
  sample_data(x)$is.neg = sample_data(x)$Sample_or_Control == "Control Sample"
  return(x)
})

# sample_data(ps)$is.neg <- sample_data(ps)$Sample_or_Control == "Control Sample"

contamdf.prev.all.blocks = lapply(psByBlock, function(y){
  isContaminant(y, method = "prevalence", neg="is.neg")
})
# contamdf.prev <- isContaminant(ps, method = "prevalence", neg="is.neg")
# table(contamdf.prev$contaminant)
```

```{r}
table(contamdf.prev.all.blocks[[1]]$contaminant)
table(contamdf.prev.all.blocks[[2]]$contaminant)
table(contamdf.prev.all.blocks[[3]]$contaminant)
```

Prevalence-based contaminant identification has identified 570, 813, and 351 contaminants than did frequency-based method.

Use different threshold
```{r}

contamdf.prev05.all.blocks = lapply(psByBlock, function(y){
  isContaminant(y, method = "prevalence", neg="is.neg", threshold = 0.5)
})
# contamdf.prev05 <- isContaminant(ps, method="prevalence", neg="is.neg", threshold = 0.5)
table(contamdf.prev05.all.blocks[[1]]$contaminant)
table(contamdf.prev05.all.blocks[[2]]$contaminant)
table(contamdf.prev05.all.blocks[[3]]$contaminant)
```

Prevalence-based contaminant identification with cutoff .05 has identified more (663, 946, and 427) contaminants.


### Check Pt\_113

Pt\_113 is in block 1
```{r}
head(which(!contamdf.prev05.all.blocks[[1]]$contaminant))
noncont = tax_table(ps)[which(!contamdf.prev05.all.blocks[[1]]$contaminant), ]
### Do we have Klebsiella_pneumoniae in block 1?
sum(noncont %in% "s__Klebsiella_pneumoniae")
```

From the clinical tests/culture, it is known that `Klebsiella_pneumoniae` is in subject Pt\_113, but this species is identified as contaminant.


# decontam on healthy flossing data


```{r}
ps <- readRDS("./Data/pshf.rds")
ps
if(dim(otu_table(ps))[1]!=ntaxa(ps)){
  otu_table(ps) <- t(otu_table(ps))
  }
```

###  Identifying batches


```{r adding_blocks}
set1 <- c("FL002","FL003","FL004","FL005")
set2 <- c("FL006","FL007","FL008","FL009","FL010","FL011")

#       the following negative controls are in set 2 only.
nctrl <- c("Water.Keck.1","Phix.1","Phix.2","Water.ctrl.1","Water.ctrl.2")

ext.num <- sample_data(ps)$SubjectName
blocks <-ifelse(ext.num %in% set1, "Set1", "Set2")

sample_data(ps)$block <- blocks
```


###  Preprocessing

Identify the species that are not present in at least one plasma sample and removed them from the phyloseq object. Label these species as contaminants. 

```{r}
ps <- prune_taxa(taxa_sums(ps) > 0, ps)
ps_plasma <- subset_samples(ps,Sample_Type %in% c("Plasma"))
prevTaxaP <- apply(otu_table(ps_plasma), 1 ,function(x){sum(x>0)})

Contaminants1 <- names(prevTaxaP)[prevTaxaP == 0]
ps <- prune_taxa(prevTaxaP > 0, ps)
ps
```

### Adding True Sample/Control Sample
```{r}
sam <- sample_data(ps) %>% data.frame()
sam <- mutate(sam, Sample_or_Control = ifelse(Sample_Type == "Plasma", "True Sample", "Control Sample"))
sample_data(ps)$Sample_or_Control = sam$Sample_or_Control
```

### Inspect library sizes
```{r}
df <- as.data.frame(sample_data(ps)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
ggplot(data = df, aes(x = Index, y = LibrarySize, color = Sample_or_Control)) + 
  geom_point()
```


Control samples have larger reads than plasma samples.


### Identify samples by blocks
```{r}
source("psBlockResults.R")
psBlockResult <- psBlockResults(ps, sampleTypeVar = "Sample_Type", caselevels = c("Plasma"), controllevel= "Control", sampleName = "SampleCode", blockVar = "block")

psByBlock <- psBlockResult[[1]]
```

### Identify Contaminants - Frequency
In this method, the distribution of the frequency of each sequence feature as a function of the input DNA concentration is used to identify contaminants.

In our phyloseq object, "Conc.ng.ul" is the sample variable that holds the concentration information:
```{r}
psByBlock = lapply(psByBlock, function(x){
  sample_data(x)$Conc.ng.ul = sample_data(x)$Conc.ng.ul + 0.001 # there is some samples with 0.00
  return(x)
})

contamdf.freq.all.blocks = lapply(psByBlock, function(x){
  isContaminant(x, method="frequency", conc = "Conc.ng.ul")
})
# contamdf.freq <- isContaminant(ps, method="frequency", conc="Conc.ng.ul")
head(contamdf.freq.all.blocks[[1]])
head(contamdf.freq.all.blocks[[2]])

```

```{r}
table(contamdf.freq.all.blocks[[1]]$contaminant)
table(contamdf.freq.all.blocks[[2]]$contaminant)

```

decontam identified only 5 species and 5 species as contaminants in block 1 and 2, respectively. 

```{r}
head(which(contamdf.freq.all.blocks[[1]]$contaminant))
head(which(contamdf.freq.all.blocks[[2]]$contaminant))

```

```{r}
plot_frequency(psByBlock[[1]], 
               taxa_names(ps)[c(1,2)], 
               conc="Conc.ng.ul") +
  xlab("DNA Concentration (PicoGreen fluorescent intensity)")
```

We know that there are more than 5 contaminants in each block, but decontam didn't identify them.

### Identify Contaminants - Prevalence

The second contaminant identification method we’ll use the “prevalence” method. In this method, the prevalence (presence/absence across samples) of each sequence feature in true positive samples is compared to the prevalence in negative controls to identify contaminants.

In our phyloseq object, "Sample_or_Control" is the sample variable that holds the negative control information. We’ll summarize that data as a logical variable, with TRUE for control samples, as that is the form required by isContaminant.

```{r}
psByBlock = lapply(psByBlock, function(x){
  sample_data(x)$is.neg = sample_data(x)$Sample_or_Control == "Control Sample"
  return(x)
})

# sample_data(ps)$is.neg <- sample_data(ps)$Sample_or_Control == "Control Sample"

contamdf.prev.all.blocks = lapply(psByBlock, function(y){
  isContaminant(y, method = "prevalence", neg="is.neg")
})
# contamdf.prev <- isContaminant(ps, method = "prevalence", neg="is.neg")
# table(contamdf.prev$contaminant)
```

```{r}
table(contamdf.prev.all.blocks[[1]]$contaminant)
table(contamdf.prev.all.blocks[[2]]$contaminant)

```

Prevalence-based contaminant identification has identified 195, and 419 contaminants than did frequency-based method.

Use different threshold (deafult threshold is .1)
```{r}

contamdf.prev05.all.blocks = lapply(psByBlock, function(y){
  isContaminant(y, method = "prevalence", neg="is.neg", threshold = 0.5)
})
# contamdf.prev05 <- isContaminant(ps, method="prevalence", neg="is.neg", threshold = 0.5)
table(contamdf.prev05.all.blocks[[1]]$contaminant)
table(contamdf.prev05.all.blocks[[2]]$contaminant)

```

Prevalence-based contaminant identification with cutoff .05 has identified more (571, 902) contaminants.


### Check FL003 subject

This subject shows some true species immediately after flossing and 3 second thereafter.
```{r}
head(which(!contamdf.prev05.all.blocks[[1]]$contaminant))
noncont = tax_table(ps)[which(!contamdf.prev05.all.blocks[[1]]$contaminant), "Species"]
head(noncont)

```

This noncontaminant list has a lot of known contaminants.

